Can dike grassland restoration profit from ideal seed mixture-substrate combinations?

### 2023-02-16 manuscript.Rmd
Update manuscript.Rmd

1.  Sowing is a well established restoration technique to overcome dispersal limitation. Though, adapted seed mixtures to certain site conditions promise a more effective restoration. This is especially important if the restored vegetation has to protect constructions like dikes. An ideal seed mixture-substrate combination would concurrently enhance erosion protection and species-rich grasslands by a fast and successful establishment of target species.
2.  An experiment with 288 plots was conducted on both expositions of a dike and vegetation was surveyed 2018-2021. We tested different sand admixtures, substrate depths, seed densities and seed mixture types (hay meadows, dry calcareous grasslands), and measured the persistence of sown species, the Favorable Conservation Status (FCS) and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a significant but slightly positive effect on all indices. However, substrate depth, seed density and seed mixture type had no significant effect on persistence or FCS.
4.  On the north exposition, hay meadow-seed mixtures developed closer to its reference than dry grasslands. Overall, the vegetation developed in the desired direction but a recovery debt remained after four years. Plots still exhibited similarities to negative references from ruderal vegetation.
5.  The south exposition was significantly worse than the north exposition which might be due to the severe drought during establishment and the addition and removal of a geotextile. This negative effect remained over the whole observation period.
6.  Synthesis and applications: Adapted seed mixtures can make restorations more efficient. Though, varying seed mixture-substrate combinations might foster biodiversity more than a desired 'ideal' combination. Furthermore, additional efforts in management or resowing due to severe legacy effects by drought might be necessary to reduce the recovery debt.

@@

Sowing as an approach to restore semi-natural grasslands has long been established [@Kiehl.2010], is also able to restore functional diversity [@Engst.2016] and is continuously developed [e.g., @Freitag.2021]. Restoration is seen as a key factor to sustain ecosystem services and biodiversity [@COP.2014; @UnitedNations.2019b]. It is necessary to successfully restore grasslands since this habitat type provide several ecosystem services and a high biodiversity while it is worldwide endangered [@Dengler.2014; @Bardgett.2021] even by restoration through unsuited afforestation [@dudley2020]. In Europe, calcareous grasslands and hay meadows are on the red list [Category 3, vulnerable, @Janssen.2016].

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts [@Freitag.2021]. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Sowing in general is a way to overcome the dispersal filter [@myers2009; @orrock2023], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

Dikes are promising sites for the restoration of species-rich grasslands [@Liebrand.1996], because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in an agricultural landscape [@Batori.2020]. Dikes have specific characteristics with strongly exposed slopes and river and land orientation [@Batori.2016] and need a reliable vegetation cover to conserve dike security [@berendse2015]. Dikes can reconcile several requirements [@teixeira2022], but the seed mixture-substrate combination must be adapted.

The aim of this study is to identify the ideal combination of substrates and seed mixtures for north and south exposed slopes in Central Europe to develop a vital and species-rich vegetation cover. We set up an experiment and tested agricultural soil with different amounts of sand admixture and different installation depths, as well as different seed mixture types and different seed densities. We expected a better development of dry grassland on the south exposition with sandy substrate and hay meadows on north exposition with less sandy substrate. Since seeds are expensive, practitioners like to know if the slopes need a higher seed density [cf. @KleberLerchbaumer.2017].

To measure the success of restoration, we chose four indices. First, we observed the persistence of the sown species, which is a key interest of sowing as a restoration approach [@Wilsey.2021]. For this, it was surveyed if the sown species are present in the respective plot in following years. Second, we calculated the Favorable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are target species for a habitat type of interest. Third, the successional distance to the reference was determined. This measure quantifies the distance of a plot to the average reference site from an ordination [@Rydgren.2019]. To complement the indices, the success of restoration or the difference from desired conditions needs to be evaluated [@Brudvig.2017b]. Therefore, the species composition of the experimental plots were compared with reference sites. We asked following question to evaluate the restoration approaches and success after four years:

>>>>

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Though, adapted seed mixtures to certain site conditions promise a more effective restoration. This is especially important if the restored vegetation has to protect constructions like dikes. An ideal seed mixture-substrate combination would concurrently improve the ecosystem function erosion protection and enhance species-rich grasslands by a fast and successful establishment of target species.
2.  An experiment with 288 plots was conducted on both expositions of a dike and was surveyed 2018--2021. We tested different sand admixtures, substrate depths, seed densities and seed mixture types (hay meadows, dry calcareous grasslands), and measured the persistence of sown species, the Favorable Conservation Status (FCS) and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a significant but slightly positive effect on all indices. Substrate depth, seed density and seed mixture type had no significant effect on persistence or FCS.
4.  On the north exposition, hay meadow-seed mixtures developed closer to its reference than dry grasslands. Overall, the vegetation developed in the desired direction but a recovery debt remained after four years. Plots still exhibited similarities to negative references from ruderal vegetation.
5.  The south exposition was significantly worse than the north exposition which might be due to the severe drought during establishment and the addition and removal of a geotextile. This initial negative effect remained over the whole observation period.
6.  Synthesis and applications: Adapted seed mixtures can make restorations more efficient like applying hay meadows on the north exposition. However, varying seed mixture-substrate combinations might foster biodiversity like beta diversity more than a desired 'ideal' combination. Furthermore, additional efforts in management might be necessary to reduce the recovery debt and resowing after a severe drought might be sensible due to strong legacy effects.

@@

Grasslands provide several ecosystem services and a high biodiversity [@Dengler.2014; @Bardgett.2021]. However, they are globally endangered [@Bardgett.2021]even by restoration through unsuited afforestation [@dudley2020]. In Europe, calcareous grasslands and hay meadows are on the red list [Category 3, vulnerable, @Janssen.2016]. Restoration is seen as a key factor to sustain ecosystem services and biodiversity [@COP.2014; @UnitedNations.2019b] and sowing is a long established restoration approach for species-rich grasslands [@Kiehl.2010]. Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts [@Freitag.2021]. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021].

Restoration ecology can increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015] by using rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory, which will finally lead to evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Sowing in general is a way to overcome the dispersal filter [@myers2009; @orrock2023], but we focus on habitat and biotic filtering which are the main assembly processes which can be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions emerged with the rise of functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020] but it is not easy to put it into practice [@merchant2022]. Substrates are artificial soil mixtures and are used in urban areas [@Bauer.2022], quarries [@chenot-lescure2022], or dikes [@Liebrand.1996].

Dikes are promising sites for the restoration of species-rich grasslands, because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in an agricultural landscape [@Batori.2020]. Dikes have specific characteristics with strongly exposed slopes and river and land orientation [@Batori.2016] and need a vegetation cover to ensure dike stability [@berendse2015]. Dikes can reconcile several requirements [@teixeira2022], which can be fostered by an adapted seed mixture-substrate combination.

The aim of this study is to identify the ideal combination of seed mixtures and substrates for north and south exposed slopes in Central Europe, which develops a vital and species-rich vegetation cover. We set up an experiment and tested agricultural soil with different amounts of sand admixture, substrate depths, seed densities, and seed mixture types. We expected a better development of dry grassland on the south exposition with shallow sandy substrates and hay meadows on the north exposition and less sandy, deep substrates. Furthermore, higher seed densities are recommended for slopes [@KleberLerchbaumer.2017], but the benefit has not been proven, yet.

To measure the success of restoration, we chose four indices. First, we observed the persistence of the sown species, which is a key interest of sowing as a restoration approach [@Wilsey.2021]. For this purpose, the sown species of a specific plot were surveyed, if they are present in that plot in the following years. Second, we calculated the Favorable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are target species for a habitat type of interest. Third, the successional distance to the reference was determined. This measure quantifies the distance of a plot to the average reference site derived from an ordination [@Rydgren.2019]. To complement these indices, the success of restoration or the difference from desired conditions was evaluated [cf. @Brudvig.2017b]. This was reached by comparing the species composition of the experimental plots with reference sites. We asked following question to evaluate the restoration approaches and the success after four years:
